---
title: "COVID trials"
author: "Fanyi Zeng"
date: "2/9/2022"
output: github_document
---

In this project, I will analyze a dataset of COVID-19 clinical trials conducted around the world. This dataset is from ClinicalTrials.gov, which is maintained by the NIH. 

### Load packages and data

```{r load, message=FALSE}
library(tidyverse)
covid <- read.csv("data/COVID clinical trials.csv")
```

### View rows and columns

There are 5783 observations and 27 variables. Let's take a look at what those variables are.

```{r head}
view(covid)
colnames(covid)
```

### Distribution by study status

It will be interesting to see the distribution of studies by status. It seems like most studies are recruiting (2805), some are completed (1025), some are not yet recruiting (1004), and a few are active but not recruiting (526).

```{r status}
covid %>%
  count(Status) %>%
  arrange(desc(n))
covid %>%
  ggplot(aes(y = Status)) +
  geom_bar()
```

### Distribution by study types

How about the distribution of studies by types?
The majority are interventional studies (3322), followed by observational studies (2427).

```{r type}
covid %>%
  count(Study.Type) %>%
  arrange(desc(n))
covid %>%
  ggplot(aes(y = Study.Type)) +
  geom_bar()
```

### Distribution by gender

There are 5567 studies that include all genders, 162 female studies, and 44 male studies. It is surprising to me that there are more female studies than male studies, because historically there were more clinical trials and medical experiments done on males, especially white males from Western countries.

```{r gender}
covid %>%
  count(Gender) %>%
  arrange(desc(n))
covid %>%
  ggplot(aes(x = Gender)) +
  geom_bar()
```

### Distribution by enrollment

The 1st quartile of enrollment (sample size) is 60, 2nd quartile is 170, 3rd is 560. To make the data more managable, I divided the studies into four groups based on their sample sizes, from smallest (A) to largest (D). For further simplification, I also created a subset that only contains interventional and observational studies, the two major types of studies.

The distribution graphs show us that again, studies that are recruiting, completed, not yet recruiting, and active but not recruiting are the dominant groups (in descending order). We can also spot some other patterns here - the suspended, withdrawn, and terminated studies are mostly intervention studies. They all have small sample sizes, which makes sense because they stopped recruiting. It might be interesting to see what other variables are related to those discontinued studies.

```{r enrollment}
summary(covid$Enrollment)
covid <- covid %>%
  mutate(Size = case_when(
    Enrollment <= 60 ~ "A",
    Enrollment > 60 & Enrollment <= 170 ~ "B",
    Enrollment > 170 & Enrollment <= 560 ~ "C",
    Enrollment > 560 ~ "D"
  ))
covid.subset <- covid %>%
  filter(Study.Type == "Interventional" | Study.Type == "Observational")
covid.subset %>%
  ggplot(aes(y = Size, fill = Status)) +
  geom_bar() +
  facet_wrap(~Study.Type)
```

### Funding sources of discontinued studies

Maybe we can investigate the funding sources of those discontinued studies... Unfortunately, the source information (such as "other") is too vague. We can only see that secondary to "other" source of studies, a number of industry studies did not make it.

```{r discontinued-studies}
covid.subset %>%
  filter(Status == "Suspended" | Status == "Withdrawn" | Status == "Terminated") %>%
  ggplot(aes(y = Status, fill = Funded.Bys)) +
  geom_bar() +
  facet_wrap(~Study.Type)
```

### The status and types of single-gender studies

Most male, intervention studies are completed.
Most female studies are still recruiting. 
There are more completed, recruiting, or "not yet recruiting" female studies than male ones.
Very few male or female studies are withdrawn, and none are suspended or terminated.
This might indicate that single-gender studies, especially female studies, are emphasized and valued by the clinical research field.

```{r gender-studies}
covid %>%
  filter(Study.Type == "Interventional" | Study.Type == "Observational") %>%
  filter(Gender == "Female" | Gender == "Male") %>%
  ggplot(aes(y = Gender, fill = Status)) +
  geom_bar() +
  facet_wrap(~Study.Type)
```

